Generalised gravitational burst generation with Generative Adversarial Networks

### 2020-09-17 Paper_draft/main.tex
more edits

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{Paper_draft/figures/training-4.png}
    \caption{Examples of the five different waveforms that were used in training the \ac{GAN} for this study. \jordan{put in the parms here too?}}
    \label{fig:training_waveforms}
\end{figure}

@@

As \acp{GAN} are trained by updating one model at the expense of the other, they can be hard to train. The goal of GAN training is to find an equilibrium between the two models, if this cannot be found then it is said that the \ac{GAN} has failed to converge. One way to diagnose \acp{GAN} during the development process is to keep track of the loss over time. Loss plots, for example, as seen in fig. \ref{fig:lossplot} can help to identify common failure modes or to check if the \ac{GAN} has indeed converged. There is currently no notion of early stopping in \acp{GAN}, instead, training is halted after convergence and by visually inspecting the generations.

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Paper_draft/figures/Screenshot 2020-09-11 at 16.52.47.png}
    \caption{Line plots of the discriminator (blue) and generator (orange) loss v iterations. Early in training the losses oscillates as both models attempt to find their equilibrium, after which, both losses vary around 0.7.}
    \label{fig:lossplot}
\end{figure}

@@

After training the networks, the generator's function is to map $G$ : $\mathbf{z},\mathbf{c} $ $\in$
$\mathbb{R}^{100}$ $\to$ $\mathbb{R}^{1024}$. The results from the
generator can be seen in \cref{fig:gen_signals}. Each plot shows the output of
the generator for random input latent space vectors $\mathbf{z}$ and each row
corresponds to one of the five class vectors $\mathbf{c}$.~\chris{yes? What
about the plot? Anything positive to say about what you're showing? Please
comment on the quality of the generated waveforms and the good varaition within
each class representative of the variation in the training data. The point is
to show the general variation and quality.}

\chris{You also need to make it clear somewhere (maybe here) that you are
generting single polarisation waveforms with normalised amplitude fixed to
unity.}

\chris{Also, how is the reader going to judge that your waveforms are good if
we don't show any examples of the training data. I recommend a single plot
showing 5 waveforms (5 panels horizontally) for reference.}

@@

% expain what we are planning to show here
%
In this section we explore the latent space formed by the generator by interpolation. We take two random
points in the latent space and linearly interpolate between them~\chris{OK, but
how do you select the latent points? Are they the same pair of points for each
class?}. These new latent space vectors can now be fed into the generator to
make predictions on~\chris{"to make predictions on" - what does that mean?}
while keeping the class vectors constant at each of the 5 \chris{fiducial?}
training classes. The full effect shown in~\cref{fig:z_interp}~\chris{not a
sentence}. We can see that each plot shows plausible waveforms suggesting that
the generator has constructed a smooth space unlike the discrete training
case~\chris{don't confuse the issue by saying that the class space is discrtee
when later we want to show the opposite}. Additionally as each class is given
the same latent points to interpolate over, we can see that the waveforms
cluster together with respect to their parameters. Visually, the sine-gaussian
and ring-down waveforms share similar frequencies and the other signals show
similar decays and starting epochs for a common latent space input. The only
exception is for the \ac{BBH} waveforms. This can be explained by the fact that
they were trained with more variety of parameters~\chris{what is more variety
of parameters? Does it really mean anything?} and consistently have their peak
amplitudes in the last quarter of the time series~\chris{do they? In the plot
the peak amplitude locations seem to range from 0.6 back to 0.5 but that's not
too different to the ring-downs, the white noise or the pulse. The
sine-Gaussian is the one that stands out as different. Can you please verify
the correctness of the plot you are referring to.}.

\chris{I think that it might be sensible to hand-pick the latent space
locations as the origin and the point $[1,1,\ldots,1]/2$. When you do random
locations we don't know what we should expect to get.}

\ac{ML} algorithms are often described as universal function approximators. In
the case of the generator it maps samples drawn from a 100 dimensional Gaussian
distribution in the latent space to its representation of the training set. As
with any function, there should be a one to one mapping from the domain and
co-domain to allow for smooth transitions across the latent space. One
advantage of using \acp{GAN} as a waveform generator is that once it is
trained, it can perform rapid generations faster than computationally expensive
algorithms \michael{e.g.?}~\chris{you can maybe reference Seb Kahn and Green
here soon. It will be on the arXiv next week I think - they did fast CBC
waveform generation.  Also Walter Del Pozzo has a paper xoming out soon on
this. Neither use GANs though, just ML}. For complicated data sets, the network
architecture must be diverse and dense enough to capture distinct variations
from the training set.  Most GANs perform well on relatively low resolution
image generations, however, higher resolutions demand larger networks and long
training times. GANs attempting to replicate complicated structures and do not
have the necessary architecture either struggle to produce results at all or
fall into the common failure mode know as mode collapse; where the generator
produces a small variety of samples or simply memorises the training set. To
test this, we perform linear interpolations in the latent and class
space.~\chris{lots of technical talk before actually saying what you are trying
to do. Reverse this. Also try to provide references for the technical talk}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/generations/z_interp_sg.png}
    \includegraphics[width=\textwidth]{figures/generations/z_interp_rd.png}
    \includegraphics[width=\textwidth]{figures/generations/z_interp_wnb.png}
    \includegraphics[width=\textwidth]{figures/generations/z_interp_blip.png}
    \includegraphics[width=\textwidth]{figures/generations/z_interp_bbh.png}
    \caption{Latent space linear interpolation, class space is
constant.~\chris{lots more discription needed. what is being plotted against
what? What is being held fixed, what is being varied.}}
    \label{fig:z_interp}
\end{figure}

@@

In order to explore the class space we keep the latent vector constant and
interpolate through the 5 classes. We construct a path between the 5 waveforms
and show that the space is populated enough to allow for transitions between
classes. Sine-Gaussian to ringdown performs well in interpolation with each
signal being a plausible burst GW. It is obvious that the GAN has clustered
these two groups during training as they share many characteristics. The other
signals have sharper transitions but still retain plausible looking waveforms.

~\chris{this is one of  the most important sections of the paper and there
is barely 1 paragrpah about it. Please beter explain what you are doing in
general when you fix the latent space location and explore the class space. The
concept of having arbitrary points in the 5-D class space should have been made
clear earlier on in the CGAN section (or elsewhere). There also isn't currently a
reference to the corresponding figure.}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/generations/interp_blip-sg.png}
    \includegraphics[width=\textwidth]{figures/generations/interp_wnb-rd.png}
    \includegraphics[width=\textwidth]{figures/generations/interp_sg-bbh.png}
    \caption{Class space linear interpolation, class space is held
constant~\chris{no it isn't}.
Top row: gaussian blip to sine-Gaussian, middle row: white noise burst to
ringdown, bottom row: sine-Gaussian to BBH.~\chris{Again, more description,
start from the basics, what is plotted against what. Is the latent space fixed
the same for each row? Add the class space vectors on each panel. We have 5
classes and so there are 10 possible pairs but you can't fit 10 rows here so
why do 3? Why not 5 or 6? }}
    \label{fig:c_interp}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{General interpolsation within the class space}

So far the analysis has focused~\chris{not really focussed, you only dedicated
1 paragraph to it in the previous subsection} on interpolating between two
classes of signals with the aim to use the interpolated signals as unmodelled
waveforms, instead, we may consider random mixtures of classes~\chris{don't
sell this according to our thought processes as we did the research, sell it as
"we just explained the pair interpolation since it's a stepping stone to the
general interpolation"}. We defined the classes in one-hot encoding framework,
that is, each class resides at the corner points of a 5 dimensional
cube~\chris{this should have already been stated n earlier sections}. In order
to generate an even mixture~\chris{by definition, it won't be an even mixture
if points are randomly sampled} of classes we sample points uniformly in this
space. \jordan{this would be the box}. Alternativly we can sampe points from
the plane that intersects the four corners, which for a 5-dimensional case is,
sample from a 4-simplex.~\chris{please dedicate an entire paragraph to the
simplex and box explanation. You may even need an equation or 2.} This can be
seen in~\ref{fig:unmodelled_samples}. \jordan{why? this enforces the vector
points to sum to one. Why is that good? Well it keeps the points lying on the
simplex.}~\chris{indeed. The reasons were originally that the plane is the
simplest structure (a hyper-surface or manifold if you;re being fancy) in the
class space that intersects all classes and has a symmetry such that no class
is favoured over any other - however that property is largely due to the
locations within the one-hot encoding space. Sampling from the simplex can be
seen as sampling from between the classes - truly interpolating within the
space. Alternatively, the cube is not the simplest space since it contains
points up to a distance of unity from the closst class e.g. [0,0,0,0,0] is a
distance of 1 away from all classes (as is [1,1,1,1,1]). This space contains
the simplex as a subspace but also allows for some level of extrapolation.
Hence we would expect the unexpected in the cube and hope for less surprises
with the simplex. THere is also the possibility of exploring outside the cube.
}

@@

% remove these comments
%
\begin{comment}
We can now think of ways to properly sample from the space between these points
in a uniform way. Uniformity enforces the condition that we have a variety of
waveforms. Thge easiest being to uniformly sample points in this space and use
them as the class vectors for the generator. Another way would be to keep the
magnitude of the class vector as 1. For a 5-dimensional space we would sample
from the 4-dimensional plane that intersects the corner points.  \end{comment}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Searching for generalised burst - a CNN analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

The CNN~\chris{whoah, hold on. What CNN? You said you were going to use one but
can we get a few more details first. what's the input, what's the output,
what's the loss function, then point us to a table of params in the appendix.}
is trained to distinguish between two classes: signals in additive Gaussian
noise and Gaussian noise, where “signals” are taken from either the
Conditional, Uniform or Simplex cases~\chris{the signals+noise and noise
statement is repetetive of the previous paragraph.}. This means in total there
are three \acp{CNN} to train and all \acp{CNN} share the same architectural
structure. All of the training data used is whitened using the Advanced LIGO
design sensitivity \ac{PSD}, such that there is equal noise power at each
frequency and the data is correctly normalised~\chris{but what is correctly
normalised. Remember trhat we have to be reproducible so you have to make sure
that someone reading this can repeat your experiment.}. This procedure is
applied to signals whose \acp{SNR} are sampled uniformly in the range $[1-16]$.
For each run~\chris{what is a run} the training data consists of $2\times 10^5$
signals which contain one half noise only and one half signal contained in
noise. Of that, 80\% is used for training and 20\% used for validation. A
different testing set is also used in the analysis that is $2\times 10^5$ in
size.~\chris{This final sentence is worded strangely but also why do we have
such a large testing set? I can only imagine that we did this so that we could
go to small $\sim 10^{-5}$ false alarm rates. Is that right?}

>>>>

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/training-sample.png}
    \caption{Examples of the five different waveforms that were used in training the \ac{GAN} for this study. \jordan{put in the parms here too?}}
    \label{fig:training_waveforms}
\end{figure}

@@

As \acp{GAN} are trained by updating one model at the expense of the other, they can be hard to train. The goal of GAN training is to find an equilibrium between the two models, if this cannot be found then it is said that the \ac{GAN} has failed to converge. One way to diagnose \acp{GAN} during the development process is to keep track of the loss and accuracy over time. Loss plots, for example, as seen in fig. \cref{fig:lossplot} can help to identify common failure modes or to check if the \ac{GAN} has indeed converged. Accuracy is the number of correct predictions made divided by total number of predictions. \jordan{mention why accuracy can be misleading? \cite{https://en.wikipedia.org/wiki/Accuracy_paradox}} There is currently no notion of early stopping in \acp{GAN}, instead, training is halted after convergence and by visually inspecting the generations.

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{figures/losses.png}
    \caption{Plot of the discriminator and generator loss and accuracy as a function of iterations. Early in training the losses oscillates as both models attempt to find their equilibrium, after which, both losses vary around 0.7-1.}
    \label{fig:lossplot}
\end{figure}

@@

After training the networks, the generator's function is to map $G$ : $\mathbf{z},\mathbf{c} $ $\in$
$\mathbb{R}^{100}$ $\to$ $\mathbb{R}^{1024}$. The results from the
generator can be seen in \cref{fig:gen_signals}. Each plot shows the output of
the generator for random input latent space vectors $\mathbf{z}$ and each row
corresponds to one of the five class vectors $\mathbf{c}$. The generations show good quality likeliness and good variation within each class representative of the variation of the training data. All generations are single plus polarised waveforms and their amplitudes are normalised to unity after generation.

@@

% expain what we are planning to show here
%
\ac{ML} algorithms are often described as universal function approximators. In
the case of the generator it maps samples drawn from a 100 dimensional Gaussian
distribution in the latent space to its representation of the training set. As
with any function, there should be a one to one mapping from the domain and
co-domain to allow for smooth transitions across the latent space. One
advantage of using \acp{GAN} as a waveform generator is that once it is
trained, it can perform rapid generations faster than computationally expensive
algorithms \michael{e.g.?}~\chris{you can maybe reference Seb Kahn and Green
here soon. It will be on the arXiv next week I think - they did fast CBC
waveform generation.  Also Walter Del Pozzo has a paper xoming out soon on
this. Neither use GANs though, just ML}. For complicated data sets, the network
architecture must be diverse and dense enough to capture distinct variations
from the training set.  Most GANs perform well on relatively low resolution
image generations, however, higher resolutions demand larger networks and long
training times. GANs attempting to replicate complicated structures and do not
have the necessary architecture either struggle to produce results at all or
fall into the common failure mode know as mode collapse; where the generator
produces a small variety of samples or simply memorises the training set. To
test this, we perform linear interpolations in the latent and class
space. We take two random
points in the 100-dimensional latent space and linearly interpolate between them. These new interpolated latent space vectors along with class vectors can now be fed into the generator to produce interpolated waveforms. \cref{fig:z_interp} shows the generated waveforms, with the class vectors held constant along each row. We can see that each plot shows plausible waveforms suggesting that
the generator has constructed a smooth traversable, space. Additionally, as each class is given
the same latent points to interpolate over, we can see that the waveforms
cluster together with respect to their parameters. Visually, the sine-Gaussian
and ring-down waveforms share similar frequencies and the other signals show
similar decays and starting epochs for a common latent space input.

\chris{I think that it might be sensible to hand-pick the latent space
locations as the origin and the point $[1,1,\ldots,1]/2$. When you do random
locations we don't know what we should expect to get.} \jordan{the origin point for some signals either produce no data or rubbish generations}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/generations/z_interp_sg.png}
    \includegraphics[width=\textwidth]{figures/generations/z_interp_rd.png}
    \includegraphics[width=\textwidth]{figures/generations/z_interp_wnb.png}
    \includegraphics[width=\textwidth]{figures/generations/z_interp_blip.png}
    \includegraphics[width=\textwidth]{figures/generations/z_interp_bbh.png}
    \caption{Generated interpolated waveforms plotted as a function of time. The latent space inputs for the first and last columns are random vectors drawn from a Gaussian distribution and the between panels are linearly interpolated vectors between these two points. Each row keeps on of the class vetors constant thoughout the latent space interpolation. On the first four rows each x axis is zoomed for ease of viewing, while the last row shows the full one second time interval.}
    \label{fig:z_interp}
\end{figure}

@@

While the \ac{GAN} is trained on distinct one-hot encoded classes, we may test arbitrary points in the 5-dimensional space to produce indistinct or "inbetween" waveforms. In order to explore the class space we keep the latent vector constant and
interpolate through the 5 classes. We construct a path between two waveforms
and show in \cref{fig:c_in	terp} that the space is populated enough to allow for transitions between
classes.  Sine-Gaussian to ringdown performs well in interpolation with each
signal being a plausible burst GW. It is obvious that the GAN has clustered
these two groups during training as they share many characteristics. The other
signals have sharper transitions but still retain plausible looking waveforms.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/generations/interp_blip-sg.png}
    \includegraphics[width=\textwidth]{figures/generations/interp_wnb-rd.png}
    \includegraphics[width=\textwidth]{figures/generations/interp_sg-bbh.png}
    \caption{Generated class interpolated waveforms as a function of time. The latent space vectors are held constant for each panel while each row shows generations using linearly interpolated classes as inputs to the generator. Top row: gaussian blip to sine-Gaussian, middle row: white noise burst to
ring-down, bottom row: sine-Gaussian to BBH.~\chris{10 possible pairs but you can't fit 10 rows here so
why do 3? Why not 5 or 6? }/\jordan{add vectors above panels.}}
    \label{fig:c_interp}
\end{figure}

\subsection{General points within the class space}
We have shown that the \acp{GAN} latent space and class space has structure that can be navigated through with interpolation between paris of signals. Taken a step further, we can sample from the class space in novel ways to create new inputs for the generator. These new points are categorised by the method used to sample from the class space. The methods we use are divided into the following:

\begin{itemize}
%
\item {\bf Conditional}: The original one-hot categorical generations from the \ac{GAN}.
These signals are the closest to the training set.
%
\item {\bf Uniform}: Generated using a uniform distribution U[0,1] as the input
class vector.
%
\item {\bf Simplex}: Generated using a Dirichlet distribution that samples from the
4-simplex as the input class vector and includes within it the conditional
locations.
%
\end{itemize}

 In order
to generate an even mixture~\chris{by definition, it won't be an even mixture
if points are randomly sampled} of classes we sample points uniformly in this
space. \jordan{this would be the box}. Alternativly we can sampe points from
the plane that intersects the four corners, which for a 5-dimensional case is,
sample from a 4-simplex.~\chris{please dedicate an entire paragraph to the
simplex and box explanation. You may even need an equation or 2.} This can be
seen in~\ref{fig:unmodelled_samples}. \jordan{why? this enforces the vector
points to sum to one. Why is that good? Well it keeps the points lying on the
simplex.}~\chris{indeed. The reasons were originally that the plane is the
simplest structure (a hyper-surface or manifold if you;re being fancy) in the
class space that intersects all classes and has a symmetry such that no class
is favoured over any other - however that property is largely due to the
locations within the one-hot encoding space. Sampling from the simplex can be
seen as sampling from between the classes - truly interpolating within the
space. Alternatively, the cube is not the simplest space since it contains
points up to a distance of unity from the closst class e.g. [0,0,0,0,0] is a
distance of 1 away from all classes (as is [1,1,1,1,1]). This space contains
the simplex as a subspace but also allows for some level of extrapolation.
Hence we would expect the unexpected in the cube and hope for less surprises
with the simplex. THere is also the possibility of exploring outside the cube.
}

@@

@@

\section{Searching for generalised burst - a CNN analysis}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/SNR8.png}
    \caption{\ac{CNN} training data. Whitened noise free sine-Gaussian timeseries as seen by Hanford (red) and Livingston (light blue) detectors. These signals with fixed SNR = 8, re-scaled by the two detector network SNR. The dark blue time series for each plot shows either the same signal with added detector noise from Hanford (left) or Livingston (right). The signals plus added noise is used to train, test and validate the \ac{CNN}.}
\end{figure}

@@

The CNN~\chris{whoah, hold on. What CNN? You said you were going to use one but
can we get a few more details first. what's the input, what's the output,
what's the loss function, then point us to a table of params in the appendix.}
is trained to distinguish between two classes: signals in additive Gaussian
noise and Gaussian noise, where â signalsâ are taken from either the
Conditional, Uniform or Simplex cases~\chris{the signals+noise and noise
statement is repetetive of the previous paragraph.}. This means in total there
are three \acp{CNN} to train and all \acp{CNN} share the same architectural
structure. All of the training data used is whitened using the Advanced LIGO
design sensitivity \ac{PSD}, such that there is equal noise power at each
frequency and the data is correctly normalised~\chris{but what is correctly
normalised. Remember trhat we have to be reproducible so you have to make sure
that someone reading this can repeat your experiment.}. This procedure is
applied to signals whose \acp{SNR} are sampled uniformly in the range $[1-16]$.
For each run~\chris{what is a run} the training data consists of $2\times 10^5$
signals which contain one half noise only and one half signal contained in
noise. Of that, 80\% is used for training and 20\% used for validation. A
different testing set is also used in the analysis that is $2\times 10^5$ in
size.~\chris{This final sentence is worded strangely but also why do we have
such a large testing set? I can only imagine that we did this so that we could
go to small $\sim 10^{-5}$ false alarm rates. Is that right?}
